Perturbing SU($n$)-symmetric interactions

### 2020-03-06 collective_sun/sun_perturbation.tex
clarified that the spectral gap occurs with long-range interactions
added section on simplifying the multi-body eigenvalue problem further in the presence of additional symmetries (such as isotropy)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with power-law interactions}
\label{sec:spectral_gap}

@@

@@

The process of constructing and simplifying diagrams to compute the
coefficients $w\p{g}$ of the diagrammatic operator product expansion
in \eqref{eq:sym_prod_group} is systematic enough to be converted into
an algorithm for execution on a computer.  The last step necessary to
automate the evaluation of \eqref{eq:sym_prod_group} is expansion of
the multi-spin operators $O\p{g}$ in a basis for the permutationally
symmetric manifold $\M_0$.  Without loss of generality, we consider a
single permutationally symmetric $M$-spin operator $Q$; our task is
essentially to find coefficients of the expansion
\begin{align}
  \P_0 Q \P_0
  = \sum_{a,b} \bk{a|Q|b} \op{a}{b},
\end{align}
where $\set{\ket{a}},\set{\ket{b}}$ are complete bases for $\M_0$, and
the choice of $M$ spins on which $Q$ acts is arbitrary.  In a system
of $N$ spins with $n$ states each, a suitable basis for $\M_0$ can be
labeled by the occupation number of each single-spin state, namely
$a=\p{a_1,a_2,\cdots,a_n}$ with each $a_j\ge 0$ and $\sum_ja_j=N$.
Written out explicitly,
\begin{align}
  \ket{a} = \f1{\sqrt{\C\p{a}}}
  \sum_{\substack{\t{distinct}\\\t{permutations}\\\Pi~\t{of}~\tilde a}}
  \Pi \ket{\tilde a},
  &&
  \C\p{a} \equiv \f{\p{\sum_j a_j}!}{\prod_ja_j!}
  &&
  \ket{\tilde a} \equiv \bigotimes_{j=1}^n \ket{j}^{\otimes a_j},
\end{align}
where the multinomial coefficient $\C\p{a}$ counts the number of
distinct ways to permute the tensor factors of $\ket{\tilde a}$, and
enforces $\bk{a|a}=1$.  We denote the set of all valid assignments of
$S$ spins to $n$ states by $\A_n\p{S}$, such that the standard basis
for the permutationally symmetric manifold of $S$ spins is
$\set{\ket{a}:a\in\A_n\p{S}}$.  Identifying this standard basis allows
us to expand
\begin{align}
  \bk{a|Q|b} = \sum_{\alpha,\beta\in\A_n\p{M}}
  \delta_{a-\alpha,b-\beta}
  \sqrt{\C\p{a-\alpha}\C\p{b-\beta}}
  \sqrt{\f{\C\p{\alpha}\C\p{\beta}}{\C\p{a}\C\p{b}}}
  \bk{\alpha|Q|\beta},
  \label{eq:multi_body_eval}
\end{align}
where the differences $a-\alpha$ and $b-\beta$ are understood to be
taken element-wise; $\delta_{a-\alpha,b-\beta}=1$ if
$a-\alpha=b-\beta$ and zero otherwise; and for ease of notation we
define $\C\p{c}\equiv 0$ if any entries of $c$ are negative.  We sum
over both $\alpha$ and $\beta$ above merely to keep the expression
symmetric with respect to transposition; in practice, one can simply
sum over $\alpha\in\A_n\p{M}$ and set $\beta\equiv b-a+\alpha$.  Note
that, by slight abuse of notation, the operator $Q$ on the left of
\eqref{eq:multi_body_eval} acts on an arbitrary choice of $M$ spins
(out of $N$), whereas the operator $Q$ on the right of
\eqref{eq:multi_body_eval} is simply an $M$-spin operator, with matrix
elements $\bk{\alpha|Q|\beta}$ evaluated with respect to the $M$-spin
states $\ket\alpha,\ket\beta$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with long-range power-law
  interactions}
\label{sec:spectral_gap}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Additional symmetries}

If the interaction Hamiltonian $H_0$ and multi-body perturbation
$\V_M$ have additional symmetries, such as isotropy or nonchirality,
then we can use this symmetry to reduce the eigenvalue problem in
\eqref{eq:H_VM_eig_simp} even further.  To this end, we define the set
$\S_{LD}\p{M}$ of equivalence classes in $\T_{LD}\p{M}$ under the
relevant symmetry.  In the case of isotropic two-body perturbations,
for example, the set of translationally-symmetrized spin choices
$\T_{LD}\p{2}$ can be classified by {\it displacements} on a lattice
(i.e.~a vector from one chosen spin to the other), and we can equate
all displacements that have the same magnitude, so elements of
$\S_{LD}\p{2}$ can be classified by all by {\it distances} between any
two points on the lattice.  For any member $s\in\S_{LD}\p{M}$, we then
define a uniform superposition $\ket{s}$ over all corresponding
members of $\T_{LD}\p{M}$, as well as the projector $\P_\S$ onto the
space of uniform superpositions:
\begin{align}
  \ket{s} \equiv \f1{\sqrt{\abs{s}}} \sum_{k\in s} \ket{k},
  &&
  \P_\S \equiv \sum_{s\in\S_{LD}\p{M}} \op{s},
\end{align}
where $\abs{s}$ is the number of members of $\T_{LD}\p{M}$ in the
equivalence class $s$.  We then expand the (symmetric) coefficient
vector
\begin{align}
  \v{\m w} = \P_\S \v{\m w}
  = \sum_{s\in\S_{LD}\p{M}} \ut{w}_s \ket{s},
  &&
  \ut{w}_s \equiv w_s \sqrt{\abs{s}},
\end{align}
as well as the symmetric projection of the matrix $\check{\m h}$:
\begin{align}
  \P_\S \check{\m h} \P_\S
  = \sum_{s\in\S_{LD}\p{M}} \ket{s} \sp{\f{h_s}{\abs{s}} \bra{s}
    + \sum_{d\in\ZZ_L^D} \f{h_{d,0}}{\sqrt{\abs{s}}} \sum_{a=1}^M
    \f1{\sqrt{\abs{s_{d;a}}}} \bra{s_{d;a}}},
\end{align}
where $h_s\equiv\sum_{p,q\in k}h_{pq}$ for any $k\in s$; and,
enforcing a canonical choice of $k\in s$ for all $s\in\S_{LD}\p{M}$,
$s_{d;a}$ is the equivalence class of the corresponding $k_{d;a}$,
with $\ket{s_{d;a}}\equiv0$ if $\ket{k_{d;a}}=0$.  The
symmetry-reduced eigenvalue problem is then
\begin{align}
  \P_\S \check{\m h} \P_\S \c \v{\m w} = \p{M h_0 + \Delta} \v{\m w}.
\end{align}

@@

The process of constructing and simplifying diagrams to compute the
coefficients $w\p{g}$ of the diagrammatic operator product expansion
in \eqref{eq:sym_prod_group} is systematic enough to be converted into
an algorithm for execution on a computer.  The last step necessary to
automate the evaluation of \eqref{eq:sym_prod_group} is expansion of
the multi-spin operators $O\p{g}$ in a basis for the permutationally
symmetric manifold $\M_0$.  Without loss of generality, we consider a
single permutationally symmetric $M$-spin operator $Q$; our task is
essentially to find coefficients of the expansion
\begin{align}
  \P_0 Q \P_0
  = \sum_{a,b} \bk{a|Q|b} \op{a}{b},
\end{align}
where $\set{\ket{a}},\set{\ket{b}}$ are complete bases for $\M_0$, and
the choice of $M$ spins on which $Q$ acts is arbitrary.  In a system
of $N$ spins with $n$ states each, a suitable basis for $\M_0$ can be
labeled by the occupation number of each single-spin state, namely
$a=\p{a_1,a_2,\cdots,a_n}$ with each $a_j\ge 0$ and $\sum_ja_j=N$.
Written out explicitly,
\begin{align}
  \ket{a} = \f1{\sqrt{\C\p{a}}}
  \sum_{\substack{\t{distinct}\\\t{permutations}\\\Pi~\t{of}~\tilde a}}
  \Pi \ket{\tilde a},
  &&
  \C\p{a} \equiv \f{\p{\sum_j a_j}!}{\prod_ja_j!}
  &&
  \ket{\tilde a} \equiv \bigotimes_{j=1}^n \ket{j}^{\otimes a_j},
\end{align}
where the multinomial coefficient $\C\p{a}$ counts the number of
distinct ways to permute the tensor factors of $\ket{\tilde a}$, and
enforces $\bk{a|a}=1$.  We denote the set of all valid assignments of
$S$ spins to $n$ states by $\A_n\p{S}$, such that the standard basis
for the permutationally symmetric manifold of $S$ spins is
$\set{\ket{a}:a\in\A_n\p{S}}$.  Identifying this standard basis allows
us to expand
\begin{align}
  \bk{a|Q|b} = \sum_{\alpha,\beta\in\A_n\p{M}}
  \delta_{a-\alpha,b-\beta}
  \sqrt{\f{\C\p{\alpha}\C\p{a-\alpha}\C\p{\beta}\C\p{b-\beta}}
    {\C\p{a}\C\p{b}}}
  \bk{\alpha|Q|\beta},
  \label{eq:multi_body_eval}
\end{align}
where the differences $a-\alpha$ and $b-\beta$ are understood to be
taken element-wise; $\delta_{a-\alpha,b-\beta}=1$ if
$a-\alpha=b-\beta$ and zero otherwise; and for ease of notation we
define $\C\p{c}\equiv 0$ if any entries of $c$ are negative.  We sum
over both $\alpha$ and $\beta$ above merely to keep the expression
symmetric with respect to transposition; in practice, one can simply
sum over $\alpha\in\A_n\p{M}$ and set $\beta\equiv b-a+\alpha$.  Note
that, by slight abuse of notation, the operator $Q$ on the left of
\eqref{eq:multi_body_eval} acts on an arbitrary choice of $M$ spins
(out of $N$), whereas the operator $Q$ on the right of
\eqref{eq:multi_body_eval} is simply an $M$-spin operator, with matrix
elements $\bk{\alpha|Q|\beta}$ evaluated with respect to the $M$-spin
states $\ket\alpha,\ket\beta$.
